Garbage Collection Evaluation Infrastructure\\ for the Cog VM

### 2018-06-20 paper.tex
Move labels to their section lines, othewrwise they don't label anything and the \ref{foo} entries don't work.
Add a reference for the Spur paper in ISMM.  Reference it at the end of the first para in Sec 3, "Reference Implementations".

\keywords{Benchmark, Garbage Collector, Virtual machine, Managed runtime}  %% \keywords are mandatory in final camera-ready submission

@@

The Cog virtual machine (VM), the default VM for multiple programming languages such Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample} and Newspeak \cite{NewspeakOopsla}, currently features in production a stop-the-world Mark-Compact algorithm as the full garbage collector (GC) algorithm. The algorithm has a high throughput but a high pause time during which the application is not responsive (the GC interrupts the application for multiple seconds on modern Macbooks starting from multiple Gbs heaps). For interactive applications, a new algorithm is required with a smaller pause time.

@@

First, we contacted multiple companies looking for use of the Cog VM in production with memory intensive applications (>1Gb heaps) and built benchmarks out of one of their deployed applications. As an example, we will discuss in the paper in Section \ref{sec:mooseBench} the Moose benchmarks implemented with the help of Feenk\footnote{https://feenk.com/}. Part of the business of Feenk consists in analysing software written in multiple programming languages using the open-source framework Moose~\cite{MoosePaper1,MooseBook1}. Using Moose, the application parses the software to analyse into a model, performs analysis on it and lastly releases the model. Models currently used are encoded in up to 11Gb. This behavior was turned into three benchmarks, growing, accessing and shrinking the heap.

@@

To evaluate new full GC algorithms, we need to compare it against existing algorithms. Two of the most common GC algorithms are Mark-Compact and Mark-Sweep. We describe here the current Mark-Compact implementation, which has been in production for the past few years, and an implementation of Mark-Sweep we introduced as a reference earlier this year.

\subsection{Mark-Compact}

@@

\subsection{Mark-Sweep}

>>>>

\keywords{Benchmark, Garbage Collector, Virtual Machine, Managed Runtime}  %% \keywords are mandatory in final camera-ready submission

@@

The Cog virtual machine (VM), the default VM for multiple programming languages such Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample} and Newspeak \cite{NewspeakOopsla}, currently features in production a stop-the-world Mark-Compact algorithm as the full garbage collector (GC) algorithm. The algorithm has a high throughput but a high pause time during which the application is not responsive (the GC interrupts the application for multiple seconds on modern Macbooks when using multiple Gbs heaps). For interactive applications, a new algorithm is required with a smaller pause time.

@@

First, we contacted multiple companies looking for production use of the Cog VM with memory intensive applications (>1Gb heaps) and built benchmarks out of one of the deployed applications. As an example, we will discuss in the paper in Section \ref{sec:mooseBench} the Moose benchmarks implemented with the help of Feenk\footnote{https://feenk.com/}. Part of the business of Feenk consists in analysing software written in multiple programming languages using the open-source framework Moose~\cite{MoosePaper1,MooseBook1}. Using Moose, the application parses the software to analyse into a model, performs analysis on it and lastly releases the model. Models currently used are encoded in up to 11Gb. This behavior was turned into three benchmarks, growing, accessing and shrinking the heap.

@@

To evaluate new full GC algorithms, we need to compare it against existing algorithms. Two of the most common GC algorithms are Mark-Compact and Mark-Sweep. We describe here the current Mark-Compact implementation, which has been in production for the past few years, and an implementation of Mark-Sweep we introduced as a reference earlier this year.  Cog is written to allow new object representations to be written and used.  Spur \cite{SpurMirandaBera} is the second object representation written for Cog.  Spur avoids the cost of an explicit read barrier to follow forwarding pointers on most accesses by handling the forwarding check on the failure side of cached dynamic message send; any send to a forwarded object will fail, and forwarding pointers in the current frame are followed on the failure side of lookup.  Hence in Spur all objects have room for a forwarding pointer.  This is key to the implementation of the Mark-Compact algorithm.

\subsection{Mark-Compact} \label{sec:refmc}

@@

\subsection{Mark-Sweep} \label{sec:refms}
